---
output:
  pdf_document: default
---

# Bootstrap Distributions Part 2

**Load required packages first**
```{r, message=FALSE}
library(resampledata)
```

## Solution Last Class: Bangladesh Arsenic Case Study



## Bangladesh Arsenic Case Study

Arsenic is a naturally occurring element in the groundwater in Bangladesh. Much of this water is used for drinking in rural areas,
so arsenic poisoning is a serious health issue. The dataset **Bangladesh** contains measurements on arsenic, chlorine, and cobalt levels (in parts per billion, ppb) present in each of 271 groundwater samples. 


a. Calculate the mean and standard deviation and size of the arsenic level of the sample.

```{r}
# Load and summarize the sample
arsenic <- Bangladesh$Arsenic
samp.mean <- mean(arsenic)
samp.mean
samp.sd <- sd(arsenic)
samp.sd
n.arsenic <- length(arsenic) # how many observations in arsenic
n.arsenic
```

b. Create a histogram to show the shape of the distribution of the sample data. How would you describe the shape?

```{r}
hist(arsenic)
# Compares sample with normal distribution
qqnorm(arsenic)
qqline(arsenic)
```

c. What are the center and spread (standard error) of the bootstrap resample?

```{r}
N <- 10^5 # Number of bootstrap samples
boot.dist <- numeric(N) #create vector to store bootstrap means

# Set up a for loop!

for (i in 1:N)
{
  x <- sample(arsenic, n.arsenic, replace = TRUE)
  boot.dist[i] <- mean(x)
}

mean(boot.dist) #Calculate center of bootstrap dist
sd(boot.dist) #Calculate bootstrap standard error

# Show bootstrap distribution
hist(boot.dist,  xlab = "xbar",
     main = "Bootstrap Distribution")

# Add a red line at the observed sample mean
abline(v = samp.mean, col = "red", lwd = 2, lty = 2)

# Add a blue line at the center of bootstrap dist s
abline(v = mean(boot.dist), col = "blue", lwd = 2, lty = 2)

# Compares bootstrap dist with normal dist
qqnorm(boot.dist)
qqline(boot.dist)
```

## Interval Estimates for Mean Arsenic Level

a. Compute a 95% bootstrap percentile confidence interval for the mean arsenic level in groundwater in Bangladesh.

```{r}
# Find cutoffs for 95% bootstrap CI
lower <- quantile(???, probs = ???)
upper <- quantile(???, probs = ???)
lower
upper
```

b. Interpret the practical meaning of the interval estimate.

```{r}
hist(boot.dist,  xlab = "xbar",
     main = "Bootstrap Distribution")

# Add a red line at the observed sample mean
abline(v = ???, col = "red", lwd = 2, lty = 2)

# Add blue lines at the 2.5 and 97.5 percentiles
abline(v = lower, col = "blue", lwd = 2, lty = 2)
abline(v = upper, col = "blue", lwd = 2, lty = 2)
```

This interval is giving a range of plausible values for the mean arsenic level of all groundwater in Bangladesh. We are 95% confident that the mean arsenic level is between ??? ppb and ??? ppb.


c. Sometimes it is nice to describe the interval as some value plus or minus some margin of error. Recall with normal distributions, approximately 95% of the data is within 2 standard deviations of center of the distribution.  Construct a symmetric 95% bootstrap confidence interval for the mean arsenic level.

```{r}
samp.mean - 2*??? #going 2 SE's below the point est
samp.mean + 2*??? #going 2 SE's below the point est
```



## North Carolina Birth Weights Revisited

a. Give a 95% bootstrap percentile confidence interval for the mean weight of babies born in North Carolina in 2004. 

```{r}
mean.weight <- ???(NCBirths2004$Weight)
sd.weight <- ???(NCBirths2004$Weight)
n.weight <- ???(NCBirths2004$Weight)

N <- 10^5 # Number of bootstrap samples
boot.weight <- numeric(N) #create vector to store bootstrap means

for (i in 1:N)
{
  y <- sample(???, ???, replace = TRUE)
  boot.weight[i] <- ????
}


# Find cutoffs for 95% bootstrap CI
lower <- quantile(???, probs = ???)
upper <- quantile(???, probs = ???)
lower
upper

hist(boot.weight,  xlab = "xbar",
     main = "Bootstrap Distribution")

# Add a red line at the observed sample mean
abline(v = ???, col = "red", lwd = 2, lty = 2)

# Add blue lines at the 2.5 and 97.5 percentiles
abline(v = lower, col = "blue", lwd = 2, lty = 2)
abline(v = upper, col = "blue", lwd = 2, lty = 2)
```


b. Give a 90% bootstrap confidence interval to estimate the mean weight of all babies born in North Carolina in 2004.


```{r}
# Find cutoffs for 90% bootstrap CI
lower.cl90 <- quantile(???, probs = ???)
upper.cl90 <- quantile(???, probs = ???)
lower.cl90
upper.cl90
```

c. Interpret the practical meaning of your 90% bootstrap percentile confidence interval.


d. When we decreased the confidence level, what happened to the confidence interval estimate?

## Cable Repair Times Case Study

What is the difference between the length of commercials on basic cable channels and on extended cable channels? The table shows the total number of minutes devoted to commercials during randomly selected half-hour periods on basic and extended cable TV channels.

Basic    | 7   | 10  | 10.6 | 10.2 | 8.6 | 7.6 | 8.2 | 10.4 | 11.0 | 8.5 |     |  
---------|-----|-----|------|------|-----|-----|-----|------|------|-----|-----|-----
Extended | 3.4 | 7.8 | 9.4  | 4.7  | 5.4 | 7.6 | 5.0 | 8.0  | 7.8  | 9.6 | 6.2 | 8.1

a. Use exploratory data analysis to compare the two groups.

```{r}
times.Basic <- c(7, 10, 10.6, 10.2, 8.6, 7.6, 8.2, 10.4, 11.0, 8.5)
times.Ext <- c(3.4, 7.8, 9.4, 4.7, 5.4, 7.6, 5.0, 8.0, 7.8, 9.6, 6.2, 8.1)

# hist(times.Basic)
# hist(times.Ext)
boxplot(times.Basic, times.Ext, 
        names = c("Basic", "Extended"))

# Comparing descriptive statistics
n.Basic <- length(times.Basic)
mean(times.Basic)
sd(times.Basic)

n.Ext <- length(times.Ext)
mean(times.Ext)
sd(times.Ext)

obs.diff <- mean(times.Basic) - mean(times.Ext)
obs.diff
```

b. Generate one possible bootstrap resample (from each sample).

```{r}
sample(???, 10, replace = ???)
sample(???, 12, replace = ???)
```

c. Give a  90% bootstrap percentile confidence interval to estimate the difference between the length of commercial times.

```{r}
N <- 10^5
boot.diff.mean <- numeric(N)

for (i in 1:N)
{
  basic.samp <- sample(???, ???, replace = ???) #resample for basic
  ext.samp <- sample(???, ???, replace = ???) #resample for extended
  boot.diff.mean[i] <- ???
}

# Bootstrap mean and bootstrap standard error
mean(???)
sd(???)

# Find cutoffs for 90% bootstrap CI
lower <- quantile(???, probs = ???)
upper <- quantile(???, probs = ???)
lower
upper

# Display bootstrap distribution
hist(boot.diff.mean)
abline(v = obs.diff, col = "red", lwd = 2, lty = 2)
abline(v = lower, col = "blue", lwd = 2, lty = 2)
abline(v = upper, col = "blue", lwd = 2, lty = 2)
```

d. Interpret the practical meaning of your interval estimate.


## Wetsuit Case Study

In the 2008 Olympics there was a lot of controversy over new swimsuits that possibly provided an unfair advantage to swimmers which led to new international rules regarding swimsuit materials and coverage.[*Statistics: Unlocking the Power of Statistics*. R. Lock, P. Lock, K. Lock, E. Lock, and D. Lock] Can a swimsuit really make a swimmer faster?

A study  tested whether wearing wetsuits influences swimming velocity. Twelve competitive swimmers swam 1500 meters at maximum speed twice each. Once wearing a wetsuit and once wearing a regular bathing suit. The order of the trials was randomized. Each time, the maximum velocity in meters/sec of the swimmer was recorded.

```{r}
# Vectors containing the times of each swimmer. Ordering is critical
wetsuit <- c(1.57, 1.47, 1.42, 1.35, 1.22, 1.75, 1.64, 1.57, 1.56, 1.53, 1.49, 1.51)
none <- c(1.49, 1.37, 1.35, 1.27, 1.12, 1.64, 1.59, 1.52, 1.50, 1.45, 1.44, 1.41)
Diff <- wetsuit - none
obs.mean.diff <- mean(Diff)
```

The mean of the differences of the sample is $\bar{x}_{\rm{diff}}=$ `r print(obs.mean.diff)` meters per second.

a. Generate one possible bootstrap resample. 

```{r}
sample(???, ???, replace = ???)
```

b. Construct a 99% bootstrap percentile confidence interval to estimate this difference.

```{r}
n <- length(Diff) # number of matched-pairs in sample

N <- 10^5
boot.diff <- numeric(N)

# Create bootstrap samples by picking n difference from original sample
# With replacement, and then find mean of those differences.
for (i in 1:N)
{
  samp.diff <- sample(???, ???, replace = ???)
  boot.diff[i] <- ???
}

# Calculate lower and upper cutoffs
# for the 99% Bootstrap CI
lower <- quantile(???, probs = ???)
upper <- quantile(???, probs = ???)
lower
upper

# Display results of bootstrap distribution
hist(???, xlab = "Difference", 
     main = "Bootstrap Distribution")
abline(v = obs.mean.diff, col = "red", lwd = 2, lty = 2)
abline(v = lower, col = "blue", lwd = 2, lty = 2)
abline(v = upper, col = "blue", lwd = 2, lty = 2)

```

c. Interpret the practical meaning of your 99% bootstrap percentile confidence interval.