---
output:
  html_document: default
---
# Confidence Interval for Means and Proportions


**Load required packages first**
```{r, message=FALSE}
library(resampledata)
```

## Question 6 Worksheet 16

Using the dataset *NCBirths2004*, give a 99% confidence interval for the mean weight (in grams) of all babies born in North Carolina in 2004.

(a) What is the point estimate?

(b) What value of $t_{\alpha/2}$ should you use?

(c) Give an approximation for the standard error of the sampling distribution for the sample mean.

(d) What is the margin of error of your 99% confidence interval?

(e) Give a 99% confidence interval based on your answers above.

(f) Construct a 99% confidence interval using the t.test command in R.

(g) Interpret the meaning of your answer in practical terms. 

## Question 8 Worksheet 16 

A study randomly assigned students to take notes either writing them by hand or using a laptop. The resulting scores of the students on a test of the material are summarized in the table below:

Group | n | $\bar{x}$ | s 
----- | - | --------- | --
By hand | 38 | 25.6 | 10.8
Laptop | 40 | 18.3 | 9.0

(a) Give a 95% confidence interval using an approximation with a $t$-distribution with degrees of freedom equal to the $n_{\rm min} -1$.

(b) Interpret the meaning of your answer. Do you believe there is a difference in the exam scores of the two groups? Explain. 

(c) If you use Welch's approximation, you get $v = 72.1368 \approx 72$. Give a 95% confidence interval using an approximation with a $t$-distribution with degrees of freedom equal to the v=72.

## Question 2 Worksheet 17

A random sample of 1,082 registered voters in the US, 29% of the people polled said they do not intend to get vaccinated for COVID-19. Construct a 95% confidence interval to estimate the proportion of all adults in the US that do not plan to get vaccinated for COVID-19.

## Question 3 Worksheet 17

Find 95% confidence interval for the proportion of all adults in the US that do not plan to get vaccinated using the **Agresti-Coull** Confidence Interval for a Proportion.

## Question 5 Worksheet 17

Find 95% confidence interval for the proportion of all adults in the US that do not plan to get vaccinated using the score confidence interval for a proportion.


```{r eval=FALSE}
lower <- (phat+z.star^2/(2*n) - 
            z.star*sqrt( (phat*(1-phat))/n + z.star^2/(4*n^2) ) )/(1+z.star^2/n)
```

Find 95% confidence interval for the proportion of all adults in the US that do not plan to get vaccinated using the score confidence interval for a proportion using the prop.test command.

## Question 6 Worksheet 17

Using the data below collected from a survey, construct a 90\% interval for the difference in the proportion of all Democrats and proportion of all Republicans that do not plan to be vaccinated.

Party | Yes, will | Yes, already | No | Unsure | Total
----- | --------- | ------------ | -- | ------ | -----
Democrat | 213 | 108 | 40 | 7 | 368
Republican | 93 | 70 | 120 | 9 | 292
Total | 306 | 178 | 160 | 16 | 660

