# Sample Homework
## Adam Spiegler

**Load required packages first**
```{r, message=FALSE}
library(resampledata)
```

1. How many flights in the dataset FlightDelays were operated by each airline?
+ Here is a table displaying the counts:
```{r}
airline <- table(FlightDelays$Carrier)
airline
```
+ Here is a barplot of the results.
```{r}
barplot(airline)
```

2. Compare the distributions of flight lengths for both airlines.
+ Below is the distribution for American Airlines.
```{r}
AA <- subset(FlightDelays, Carrier == "AA")
hist(AA$FlightLength)
```

+ Below is the distribution for United Airlines.
```{r}
UA <- subset(FlightDelays, Carrier == "UA")
hist(UA$FlightLength)
```

+ The mean flight length of AA is `r mean(AA$FlightLength)`.
+ The mean flight length of UA is **`r mean(UA$FlightLength)`**.

# Useful References
* **R Markdown Cheatsheet:** (https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
* **R Markdown The Definitive Guide:** (https://bookdown.org/yihui/rmarkdown/)
* **Chunk Code Options:** (https://yihui.org/knitr/options/)

# The End